---
output:
  html_document: default
  pdf_document: default
---
White wine quality by Sumit Kumar Mahato
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#install.packages("ggplot2")
#install.packages("knitr")
#install.packages("dplyr")
#install.packages("BBmisc")
#install.packages("GGally")


#library(ggplot2)
#library(knitr)
#library(dplyr)
#library(BBmisc)
#library(GGally)

```


```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("C:/Users/sumit/Desktop/Courses/Data_analyst_nanodegree_Udacity/DAND-2/Part-2/Project")

```

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}
#wine data
wd<-read.csv("wineQualityWhites.csv")

#structure of dataset
str(wd)

#summary of the dataset
summary(wd)


#normalizing the wine dataframe

n_wd<- normalize(wd[,c(2:12)], method = "standardize", range = c(0, 5), margin = 1L, on.constant = "quiet")

#boxplots of the physicochemical features
ggplot(stack(n_wd), aes(x = ind, y = values)) +
  geom_boxplot()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

#boxplots of the physicochemical features with ylim = -5 to 5

ggplot(stack(n_wd), aes(x = ind, y = values)) +
  geom_boxplot()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))+
      scale_y_continuous(breaks = seq(-2,5,0.5), limits = c(-2,5))


```

# Univariate Analysis
The guiding question of the analysis is which physicochemicals have significant impact on the quality of the wines.

### What is the structure of your dataset?

The dataset consists of 13 variables and approximately 5000 observations.All the variables are either int or num datatype. 

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in the dataset is the quality of wines.


### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
11 physicochemical features are responsible for the quality of the wine. I am guessing that the features which are responsible for the taste, smell or after-effect of a wine could impact the quality score of the wine. Volatile acidity, citric acid, residual sugar, free sulfur dioxide, and alcohol are the features in the textfile on the dataset. In
the next section the correlation of every feature against quality is
estimated.

### Did you create any new variables from existing variables in the dataset?
No.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

I started the univariate analysis by plotting histograms of every
feature (can be found in the appendix section). For most of the features
the distribution looked approximately normal with some outliers. It was
found that boxplots could be better to compare the features with each
other. However, the scales of the features were different, therefore, the dataset was first normalized within the range of 0-1 then boxplots were created.
It can be noted that the chlorides had the lowest spread of data and 
alcohol has no outliers (is it regulated?)



# Bivariate Plots Section
```{r echo=FALSE, message=FALSE}
#the relationship between the variables are plotted against each other
#using the GGally package. The inferences from the charts are discussed
#in the following questions.


#correlation grid
theme_set(theme_minimal(20))

#set seed for reproducible results
set.seed(1610)
wd_subset<- wd[,c(2:13)]
names(wd_subset)
ggpairs(wd_subset[sample.int(nrow(wd_subset),1000),])



```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

ggpairs function was used to plot the relationship among the different
features.

Following observations were made from the above plot:

i. We could see that there is a strong positive relationship (0.842) b/w density and residual sugar which is natural as the higher the amount of residual sugar higher would be the density of the wine.

ii. Another strong relationship noted between the alcohol and
density is a negative one (-0.792). The reason could be the lower 
density of alcohol as compared to water so higher the alcohol content
in a wine lower would be the density of the wine.

iii. The total sulphur dioxide and the free sulphur dioxide has a moderate corelation of 0.6 which is justified as the latter is a constituent of the former.

iv.Apart from the above strong relationships there are couple of moderate correlation between pH and fixed acidity(-0.401) and alcohol and residual sugar(-0.456). We could assert the obvious fact that higher the acidity, lower is the pH value however, for the correlation between
alcohol and residual sugar it could be said that both are consituents
of wine and increase in one would lead to decrease in another and
vice-versa.

V. Most of the features are negatively correlated to the feature of 
interest i.e., quality with only exception of pH, sulphate, and alcohol.

vi. Alcohol content had a positive effect on the quality of wine(0.467) which was expected however, the negative correlation of density and quality was interesting which implied dense wines are less preferred.
It would be interesting to note the impact of the interacting features on quality which would be carried out in the next section.

### What was the strongest relationship you found?

The strongest relationship is found between density and residual sugar (0.842).



# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

#relationship of 2 features, 1 postiive feature with 1 negative, with
#quality of wine. Also a linear model was fit to the scatterplots


ggplot(aes(x=volatile.acidity, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of volatile acidity and alcohol \n on quality")

ggplot(aes(x=residual.sugar, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of residual sugar and alcohol \n on quality")


ggplot(aes(x=chlorides, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of chlorides and alcohol \n on quality")


ggplot(aes(x=total.sulfur.dioxide, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of total sulfur dioxide and alcohol \n 
              on quality")
ggplot(aes(x=density, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of density and alcohol \n on quality")


# 2 positive features affecting quality of wine
ggplot(aes(x=pH, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of pH and alcohol \n on quality")

ggplot(aes(x=sulphates, y= alcohol), data=wd)+
      geom_point()+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of sulphates and alcohol \n on quality")

#3 positive features affecting quality of wine
ggplot(aes(x=sulphates, y= alcohol), data=wd)+
      geom_point(aes(color=pH))+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of sulphates, pH and alcohol \n on quality")
      

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

Almost all the high quality wine i.e., with a rating 9 had alcohol
content greater than 12. Similarly, almost all the wines with poor
rating i.e., 3 had an alcohol content less than 12. Alcohol contnt strengthend almost all the features which had a negative correlation with quality.

### Were there any interesting or surprising interactions between features?
Since alcohol is the parameter which would push a wine towards a better
quality rating therefore, we can expect a positive slope linear fit line
as noted in plot of alcohol vs volatile acidity. It means that higher 
the amount of less preferred  chemical higher the alcohol needed in 
order to retain a high rating. However, in the following plots a negative slope is observed implying even though the amount of alcohol
decreased and the unwanted chemical composition increased, the quality
rating didn't drop.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

I tried to fit a linear model into the dataset. The outliers tend to 
sway the best fit line.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
#boxplots of the physicochemical features with ylim = -5 to 5

ggplot(stack(n_wd), aes(x = ind, y = values)) +
  geom_boxplot()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))+
      scale_y_continuous(breaks = seq(-2,5,0.5), limits = c(-2,5))+
      ggtitle("spread of physicochemical features")

```

### Description One
The spread of different features on a normalized scale compares the 
spread od data relative to each other. The plot also shows the presence
of outliers.

### Plot Two
```{r echo=FALSE, message=FALSE}
#correlation grid
theme_set(theme_minimal(20))

#set seed for reproducible results
set.seed(1610)
wd_subset<- wd[,c(2:13)]
names(wd_subset)
ggpairs(wd_subset[sample.int(nrow(wd_subset),1000),])
```

### Description Two
Contains all the relationship among the features in the dataset. It is
really helpful in directing the focus of the analysis by pinpointing
the features which have high impact on the feature of interest.

### Plot Three
```{r echo=FALSE, Plot_Three}
#3 positive features affecting quality of wine
ggplot(aes(x=sulphates, y= alcohol), data=wd)+
      geom_point(aes(color=pH))+
      facet_wrap(~quality, ncol=3)+
      geom_smooth(method = 'lm', color='red')+
      ggtitle("Effect of sulphates, pH and alcohol \n on quality")
```

### Description Three
The above plot shows the impact of 3 features on the feature of
interest. It is worthwhile to note that all the plots have a downward 
slope which signifies that decrease in amount of feature is balanced 
by increasein other feature to retain the quality of wine.


------

# Reflection
From the study it could be concluded that alcohol content is the single
most important feature affecting the quality of wine. A number of trials
were conducted to reach a particular plot. Such as the histograms of
each feature which was changed to a single boxplot of normalized data.
Faceting of qaulity w.r.t. various features in multivriate analysis.


###Appendix
Histograms of input features 
```{r echo=FALSE, Extra_Plots}
#distribution of volatile acidity

ggplot(aes(x=volatile.acidity), data= wd)+
      geom_histogram()


#distribution of citric acid

ggplot(aes(x=citric.acid), data= wd)+
      geom_histogram()


#distribution of residual sugar

ggplot(aes(x=residual.sugar), data= wd)+
      geom_histogram()


#distribution of chlorides

ggplot(aes(x=chlorides), data= wd)+
      geom_histogram()

#distribution of free sulphur dioxide

ggplot(aes(x=free.sulfur.dioxide), data= wd)+
      geom_histogram()


#distribution of total sulphur dioxide

ggplot(aes(x=total.sulfur.dioxide), data= wd)+
      geom_histogram()

#distribution of density

ggplot(aes(x=density), data= wd)+
      geom_histogram()

#distribution of pH
ggplot(aes(x=pH), data= wd)+
      geom_histogram()

#distribution of sulphates

ggplot(aes(x=sulphates), data= wd)+
      geom_histogram()     


#distribution of alcohol percentage

ggplot(aes(x=alcohol), data= wd)+ 
      geom_histogram()


#these plots were not helpful in finding affect of 2 or more features on
#quality
ggplot(aes(x=citric.acid, y= alcohol), data=wd)+
      geom_point(aes(color=quality))

ggplot(aes(x=residual.sugar, y= alcohol), data=wd)+
      geom_point(aes(color=quality))

ggplot(aes(x=volatile.acidity, y= alcohol), data=wd)+
      geom_point(aes(color=quality))

ggplot(aes(x=sulphates, y= alcohol), data=wd)+
      geom_point(aes(color=quality))


```

library(dplyr)
library(BBmisc)
library(GGally)